Motion planning for assessing accessibility of protein tunnels

### 2017-08-23 paper.tex
intro

\newcommand{\red}[1]{\textcolor{red}{#1}}
%\newcommand{\tylde}{\raisebox{-0.85ex}{\textasciitilde}} %does not work for vojta
\newcommand{\tylde}{$\sim$}

@@

\title{    Hey we need a title}

\author{ Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic,
{\tt vonasek@labe.felk.cvut.cz}}
\thanks{$^{2}$        
Faculty of Informatics,   Masaryk University, 
Botanick\'a 68a, 602~00 Brno, Czech Republic,\\
The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.}
}

@@

\begin{abstract}
%Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
Molecular tunnels connect catalytic active sites with the solvent and they play an important role in catalytic and thermodynamic properties.
The geometrical and physico-chemical properties of tunnels need to be determined in order to understand the transport of a ligand from and into the protein active site. 
Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamics (MD) simulations.
The tunnels are usually computed using Voronoi diagrams and therefore they can show possible movements only for a single atom.
Ligand movements inside protein can be computed using MD simulations, but they are too slow for a daily routine.
The tunnels computed for a single atom can however be used as a hint for ligand transport from or to the active site.
The task of ligand transportation through a tunnel considering its shape and also conformation changes can be formulated
 as a motion planning problem.
We propose to evaluate the tunnels based on ligand trajectories computed by motion planning, namely using
Rapidly Exploring Random Tree (RRT).
We present modifications to RRT in order to compute the trajectories around a given tunnel and also to consider the flexibility of the ligand.
Besides, methods for visualization of the trajectories and their properties are presented.
The goal of the paper is to show that motion planning can provide additional insight about tunnel properties, being
less computationally demanding than the MD simulations.
The proposed approach can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
\end{abstract}

@@

\begin{figure}[b]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3} &
\hbox{
\vbox{
\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
} 
}
\\
Protein 1CQW & Active site & Detected tunnels & Example of a  \\
             &            & (orange)          & trajectory
\end{tabular}
}
\caption{\label{fig::motiv}
    Tunnels in haloalkane dehalogenase with a possible trajectory of 1-chlorpropan ligand.
}
\end{figure}

@@

Traditionally, tunnels are detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming
%The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3}.  %or a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
 a spherical ligand (probe) and they are represented as a sequence of spheres.
Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013} or its activity of the protein towards other molecules~\cite{Pavlova2009}.
%The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
%This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
%ligand from the outside environment to the active site or vice versa.

Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.
To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.

We propose to analyze the traversability of the ligands in the tunnels using motion planning.
The principle of Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} planner is used and further modified.
The main idea of the proposed planner is to generate the random samples around a virtual sphere moving through the tunnel, which
guides the search in the configuration space.
This allows us to focus the sampling of the configuration space around the tunnel.
The flexibility of the ligand is modeled using a predefined set of typical conformations.
To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
%This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.

The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
%Besides trajectory computation, visualization of the results in important. %we also propose methods for their visualization.
Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources required for the virtual screening.
%We further propose visualization techniques for presenting the results to biochemists.
%Therefore, methods for visualization of the results  are discussed.

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\tylde}{$\sim$}

@@

\title{Motion planning for assessing accessibility of protein tunnels}

\author{Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic,
{\tt vonasek@labe.felk.cvut.cz}}
\thanks{$^{2}$        
Faculty of Informatics,   Masaryk University, 
Botanick\'a 68a, 602~00 Brno, Czech Republic,\\
The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.}
}

@@

\begin{abstract}
Chemical interactions of proteins and other molecules takes place in actives sites that are usually deeply burid inside the proteins.
Motion planning can be used to analyze accessibility of tunnels by small molecules (ligands), which can help chemists to ..
The ligands are not rigid structures, but they are rather flexible.
In this paper, we present sampling-based motion planinng method to compute trajectories of the ligands inside know protein tunnels.
The tunnels are represented by a set of spheres and they are used to focus sampling of the configuration space.
To decrease the number of degrees of freedom, the ligand flexibility is modeled by a set of predefined conformations.
We propose a new metric to measrue distances between the configurations which considers the shape of the ligands.
\end{abstract}

@@

\begin{figure}[b]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3}  \\
%\hbox{
%\vbox{
%\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
%\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
%} 
%}
Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
             &            & (orange)          %& trajectory
\end{tabular}
}
\caption{\label{fig::motiv}
    Tunnels in haloalkane dehalogenase with a possible trajectory of 1-chlorpropan ligand.
}
\end{figure}

@@

Traditionally, tunnels are detected using Voronoi diagrams~\cite{yaffe2008,caver3} assuming 
a spherical ligand (probe) and they are represented as a sequence of spheres.
Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013}. % or its activity of the protein towards other molecules~\cite{Pavlova2009}.
%The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
%This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
%ligand from the outside environment to the active site or vice versa. 
Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.

To provide biochemists  a better insight into the behavior of non-spherical ligands, it is necessary to compute the trajectories of the ligand considering its shape and possibly also its conformation changes.
This can be solved using motion planning, where the ligand is considered as a robot moving amongst protein's atoms.
Sampling-based motion planning methods are suitable candidates for this task, as they can consider non-spherical shape of the ligand as well as the conformational changes.

Proteins are dense structures which limits movements of the ligand insider the tunnels, which leads to the narrow passage problem.
To cope with this problem, we propose to utilize the principle of guided sampling~\cite{vonasek2009rrt,denny2016dynamic}.
The random samples are not generated from the whole configuration space, but mainly around the tunnel.
To enable ligand movements in narrow tunnels, a scaled-down version of the ligand is used, similarly, e.g., to~\cite{cortes2005path}.
%We propose to analyze the traversability of the ligands in the tunnels using motion planning.
The flexibility of the ligand is modeled using a predefined set of typical conformations.
%It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
%This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.

%The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
%Besides trajectory computation, visualization of the results in important. %we also propose methods for their visualization.
%Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
%In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
%The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources required for the virtual screening.
%We further propose visualization techniques for presenting the results to biochemists.
%Therefore, methods for visualization of the results  are discussed.
